The very dependent recursive structure of iterated parametricity in indexed form

### 2025-12-29 paper/paper.tex
paper: prose progress

The internalisation of parametricity in unary indexed form was studied by \cite{Moulin,Bernardy} directly leading to a type theory of iterated parametricity. Such type theory could be interpreted in a ``unary'' form of cubical set~\cite{bernardy15}, thus implicitly in fibered form since, in a binary cubical set (resp. in a unary cubical set and simplicial set as well), a set of segments over two points, that is a relevant relation, is given by first and second projection rather than as a function to $\Set$ (resp. a family is given as a \emph{total space} projected over a set of points $A:\Set$ rather than a family of type $A \rightarrow \Set$. This connection between iterated binary parametricity in fibered form and cubical sets includes an identification of degeneracies and reflexivity proofs, which can also be found in \cite{johann17}. In particular, it became clear at this point that iterated binary parametricity (without reflexivity) produces semi-cubical sets while it was also observed\footnote{The authors learned it from Hugo Moeneclaey and Thorsten Altenkirch.} that iterated unary parametricity produces augmented semi-simplicial sets.

\begin{tabularx}{\linewidth}{cYcYYl}
  \toprule
  arity                       & formalisation style        & iteration                       & relevance        & logical structure               \\
  \midrule
  \multirow{6}{0.6cm}{unary}  & \multirow{3}{1cm}{indexed} & \multirow{2}{2cm}{non-iterated} & proof-irrelevant & predicate                       \\
                              &                            &                                 & relevant         & family                          \\
                              &                            & iterated                        & relevant         & \emph{this work}                \\
                              & \multirow{3}{1cm}{fibered} & \multirow{2}{2cm}{non-iterated} & proof-irrelevant & subobject                       \\
                              &                            &                                 & relevant         & fibration                       \\
                              &                            & iterated                        & relevant         & augmented semi-simplicial set   \\
  \midrule
  \multirow{6}{0.6cm}{binary} & \multirow{3}{1cm}{indexed} & \multirow{2}{2cm}{non-iterated} & proof-irrelevant & relation                        \\
                              &                            &                                 & relevant         & correspondence                  \\
                              &                            & iterated                        & relevant         & \emph{this work}                \\
                              & \multirow{3}{1cm}{fibered} & \multirow{2}{2cm}{non-iterated} & proof-irrelevant & fibered relation ?              \\
                              &                            &                                 & relevant         & span                          & \\
                              &                            & iterated                        & relevant         & semi-cubical set                \\
  \bottomrule
\end{tabularx}

In~\citep{hr25}, this correspondence was exploited to construct at the same time augmented semi-simplicial and semi-cubical sets in \emph{indexed form} as a special case of iterated parametricity. That is, instead of the following illustration of the semi-cubical case (upto faces identities),

@@

For $\nu$ an arity, that is $1$ in the case of predicates, families and augmented semi-simplicial sets or $2$ in the case of relations, correspondences and semi-cubical sets, such sequence of was called a $\nu$-set.
Equations specifying such sequence of relevant relations were given in \cite{hr25}. Because of the dependency of $E_{n+1}$ into all $E_p$ for $p \leq n$, the family could not be expressed as a function. Instead, it was expressed as a stream whose $(n+1)$-th element $E_{n+1}$ was dependent on the bundle of all previous elements of the stream.

A $\nu$-set in indexed form is a sequence of families of $\U$, that is $\U[m]$ for some universe level $m$. We call such sequence a $\nu$-set at level $m$, whose type thus lives in $\U[m+1]$.

Table~\ref{tab:coind} describes the type of a $\nu$-set at level $m$ as a dependent stream of type families representing the limit of $n$-truncated $\nu$-sets: the recursive equation $\Xfrom{m}{n}{D} \,\defeq\, \Sigma R: \Xcomp{m}{n}[D=\D].\, \Xfrom{m}{n+1}[D=\pair{D}{R}]$ from the table formally corresponds to the stream $Stream_{\Sigma n.\, \Xto{m}{n},\, \lambda (n, \D).\, \Xcomp{m}{n}[D=\D],\,\lambda ((n, D),R).(n+1,(\pair{D}{R}))}(n, \D)$. Therefore, $\Xfrom{m}{n}$ denotes an infinite sequence $X_{n}, X_{n+1}, \ldots$ dependent on a $(<n)$-truncated $\nu$-set, $\Xto{m}{n}$, so that, when $n$ is $0$, it denotes a full $\nu$-set, written $\Xp{m}$. This is made possible because the $(<0)$-truncated $\nu$-set, $\Xto{m}{0}$, is degenerated: it is an empty family, and there is thus only one $(<0)$-truncated $\nu$-set, namely the canonical inhabitant $\kstar$ of $\unittype$.

The definition of the type of a $n$-truncated $\nu$-set is in turn described in table~\ref{tab:core}. In the infinite sequence of type families representing a $\nu$-set, the $n$-th component is a type dependent over a $\fullframe$. It is recursively defined in table~\ref{tab:frames}, using the auxiliary definitions of $\framep$, $\layer$ and $\painting$. A $\fullframe$ describes a boundary of a standard form (simplex, cube), which we decompose into $\layer$, and a $\painting$ corresponds to a filled frame. Notice that the type $\layer$ relies on an operator of frame restriction $\restrf$ which is defined in table~\ref{tab:faces}, and this restriction operator is in turn defined using auxiliary definitions $\restrl$ and $\restrp$.

Notably, the definition of $\restrl$ relies on an equality expressing the commutation of the composition of two $\restrf$. The proof of this commutation is worth being made explicit, which we do in table~\ref{tab:coh} using proof-term notations. The proof requires an induction on the dimension and on the structure of $\framep$, $\layer$, and $\painting$. This is what $\cohf$ does using auxiliary proofs $\cohl$ and $\cohp$. Even though it looks independent of the definitions from the other tables, $\cohf$ has to be proved mutually with the definitions of $\framep$, $\layer$, $\painting$, and their corresponding restrictions. More precisely, for a fixed $n$, the block of $\framep$, $\restrf$, and $\cohf$ has to be defined in one go by induction on $p$. Also, each of $\painting$, $\restrp$, and $\cohp$ is built by induction from $p$ to $n$. The $\painting$ block at $n$ relies on the $\framep$ block at $n$, but the converse dependency is only on lower $n$, so this is well-founded. Note that $\layer$, $\restrl$ and $\cohl$ are just abbreviations.

Note that for a fixed constant $n$, relying on the evaluation rules of type theory, the coherence conditions degenerate to a reflexivity proof, so that the construction builds an effective sequence of types not mentioning coherences anymore. \\

>>>>

The internalisation of parametricity in unary indexed form was studied by \cite{bernardy12} directly leading to a type theory of iterated parametricity. Such type theory could be interpreted in a ``unary'' form of cubical set~\citep{bernardy15}, thus implicitly in fibered form since, in a binary cubical set (resp. in a unary cubical set and simplicial set as well), a set of segments over two points, that is a relevant relation, is given by first and second projection rather than as a function to $\Set$ (resp. a family is given as a \emph{total space} projected over a set of points $A:\Set$ rather than a family of type $A \rightarrow \Set$. This connection between iterated binary parametricity in fibered form and cubical sets includes an identification of degeneracies and reflexivity proofs, which can also be found in \cite{johann17}. In particular, it became clear at this point that iterated binary parametricity (without reflexivity) produces semi-cubical sets while it was also observed\footnote{The authors learned it from Hugo Moeneclaey and Thorsten Altenkirch.} that iterated unary parametricity produces augmented semi-simplicial sets.

\begin{tabularx}{0.94\linewidth}{cYcYYl}
  \toprule
  arity                       & formalisation style        & iteration                       & relevance        & logical structure               \\
  \midrule
  \multirow{6}{0.4cm}{unary}  & \multirow{3}{1cm}{indexed} & \multirow{2}{2cm}{non-iterated} & proof-irrelevant & predicate                       \\
                              &                            &                                 & relevant         & family                          \\
                              &                            & iterated                        & relevant         & \emph{this work}                \\
                              & \multirow{3}{1cm}{fibered} & \multirow{2}{2cm}{non-iterated} & proof-irrelevant & subobject                       \\
                              &                            &                                 & relevant         & fibration                       \\
                              &                            & iterated                        & relevant         & augmented semi-simplicial set   \\
  \midrule
  \multirow{6}{0.4cm}{binary} & \multirow{3}{1cm}{indexed} & \multirow{2}{2cm}{non-iterated} & proof-irrelevant & relation                        \\
                              &                            &                                 & relevant         & correspondence                  \\
                              &                            & iterated                        & relevant         & \emph{this work}                \\
                              & \multirow{3}{1cm}{fibered} & \multirow{2}{2cm}{non-iterated} & proof-irrelevant & fibered relation                \\
                              &                            &                                 & relevant         & span                          & \\
                              &                            & iterated                        & relevant         & semi-cubical set                \\
  \bottomrule
\end{tabularx}

In \cite{hr25}, this correspondence was exploited to construct at the same time augmented semi-simplicial and semi-cubical sets in \emph{indexed form} as a special case of iterated parametricity. That is, instead of the following illustration of the semi-cubical case (upto faces identities),

@@

For $\nu$ an arity, that is $1$ in the case of predicates, families and augmented semi-simplicial sets or $2$ in the case of relations, correspondences and semi-cubical sets, such sequence of was called a $\nu$-set. Equations specifying $\nu$-sets were given in \cite{hr25}. Because of the dependency of $E_{n+1}$ into all $E_p$ for $p \leq n$, the family could not be expressed as a function. Instead, it was expressed as a stream whose $(n+1)$-th element $E_{n+1}$ was dependent on the bundle of all previous elements of the stream. Such a bundle was called $\Xto{m}{n}$ while the residual stream of all $E_{n}$, $E_{n+1}$, ..., all of them precisely dependent on the bundle $E:\Xto{m}{n}$ of previous $E_0$, ..., $E_{n-1}$, was called $\Xfrom{m}{n}{D}$. The type of $E_n$, also dependent on a bundle $D:\Xto{m}{n}$, was denoted $\Xcomp{m}{n}[D=\D]$, as shown on tables~\ref{tab:coind} and~\ref{tab:core}. When $n$ is $0$, $\Xto{m}{0}$ is degenerated, that is reduced to the canonical inhabitant $\kstar$ of $\unittype$. Then, $\Xfrom{m}{0}{\kstar}$ denotes the full stream $E_0, E_1, \ldots$.

The type $\Xcomp{m}{n}[D=E_0,\ldots,E_{n-1}]$ of $E_n$ represents a family whose type can be presented into the form $\fullframe(E_0,\ldots,E_{n-1}) \rightarrow \U$ where $\fullframe$ collects all the components on which $E_n$ is expected to depend. It is recursively defined in table~\ref{tab:frames}, using the auxiliary definitions of $\framep$, $\layer$ and $\painting$. A $\fullframe$ describes a boundary of a standard form (simplex, cube), which we decompose into $\layer$, and a $\painting$ corresponds to a filled frame. Notice that the type $\layer$ relies on an operator of frame restriction $\restrf$ which is defined in table~\ref{tab:faces}, and this restriction operator is in turn defined using auxiliary definitions $\restrl$ and $\restrp$.

Notably, the definition of $\restrl$ relies on an equality expressing the commutation of the composition of two $\restrf$. The proof of this commutation is worth being made explicit, which we do in table~\ref{tab:coh} using proof-term notations. The proof is by recursion on the structure of $\framep$, $\layer$, and $\painting$. This is what $\cohf$ does using auxiliary proofs $\cohl$ and $\cohp$. Note that for a fixed constant $n$, relying on the evaluation rules of type theory, the coherence conditions degenerate to a reflexivity proof, so that the construction builds an effective sequence of types not mentioning coherences anymore.

TODO: say that the table are relative to ETT and that many type coercions are implicit.

Due to the intricate dependencies, the reason the whole construction is recursively well-defined is not simple. The formalisation in \cite{hr25} was recursively building a complex structure involving $3$ levels of the construction at once. In the present paper, we untangle the previous construction, making the dependencies explicit, and the new construction lends itself to a simple and complete explanation.
